Do syntactic and semantic similarity lead to interference effects?\\ Evidence from self-paced reading and event-related potentials using German

### 2024-11-27 paper/revision2/ResponseToReviewers.tex

\noindent \textbf{Prof.\ Christopher Jarrold}\\
\noindent \textbf{Associate Editor}\\
\noindent \textbf{Journal of Memory and Language}
\vskip 1em

\noindent \textbf{Response letter for Submission JML-24-17, 2nd round of revisions}
\vskip 2em
\noindent Dear Prof. Jarrold, 
\vskip 1em
with my co-authors Shravan Vasishth and Himanshu Yadav, I would like to submit the second revision of our article titled ``Do syntactic and semantic similarity lead to interference effects? Evidence from self-paced reading and event-related potentials using German''.

We are grateful for the thoughtful and constructive comments and suggestions from yourself and the reviewers, and feel that the quality of the manuscript has been greatly improved by the revisions.
To address the concern that we overstate the implications of our findings, we have clarified at various places in the abstract and discussion of our results that our findings only apply to the present design.

Clarifying text has been added in this second round of revisions (e.g., regarding the implications, the prolonged reading times starting at the distractor and the use of syntactic and semantic cues), and we have shortened the manuscript length by three pages. We have removed redundant text passages (e.g., in the description of the SPR results).

@@

\subsubsection*{Response to E.3}
Yes, we changed all analyses in the first revision. As we had explained in the response to reviewers on August 29, 2024 (specifically in the response letter and our responses to the comments E.2 and R3.5), the priors were changed from directional priors in the original version to symmetrical priors in the revised version. This was necessary because we added trial id as a predictor as was suggested by a reviewer. There was no a priori hypothesis of the sign of the effect of trial id (do the participants get faster or slower over the course of the experiment?), therefore the prior for trial id needed to be symmetrical (positive and negative values). It is not possible to include symmetrical and directional priors within a single brms model \citep{brms}, thus all priors were changed to symmetrical ones.

Additionally, we changed the method with which the Bayes factors were computed. Initially, we had used bridge sampling, but in the revision, we switched to the Savage-Dickey method because only using that method we were able to compare models with random slopes (this was also explained in the previous response letter, Response to R1.7). These changes did not change the results qualitatively, but led to some numerical changes which probably piqued the reviewer's interest.

\subsection*{Comment E.4}
\begin{quote}
``and \textbf{their final point about the length of the current version of the manuscript}.''
\end{quote}

\subsubsection*{Response to E.4}
We have removed repetitions in the description of the statistical analyses and have shortened the description of the self-paced reading results (and the comparison to the previous studies). This reduced the length of the manuscript by three pages; after the first round of revisions the main text ended on page 77, now it ends on page 74.

@@

\subsubsection*{Response to R2.1}
We agree with the reviewer that the presentation and discussion of the SPR results was misleading and have revised it accordingly. We have revised Figure 3 to remove the focus on the critical region (see below). At the beginning of the Results section, we now anticipate our interpretation of the results and state that we believe that the distractors led to encoding interference. We also state that the results at the critical (and spill-over) region are potentially confounded because it is not possible to disentangle the long-lasting encoding effect from retrieval effects potentially arising in the critical region.

\setcounter{figure}{2}

\begin{sidewaysfigure}[h]
    \caption{Self-paced reading times with 95\% confidence intervals. Panel A and B show the pooled reading times across the whole sentence; separately for high (A) and low (B) syntactic interference due to differing sentence structure. \textcolor{blue}{Panel C shows the reading times of the sentence focusing on the regions between the distractor and critical verb for all conditions.}}
    \label{fig:whole_sentence}
    \centering
    \includegraphics[width=\textwidth]{Pandora_all_wholesentence_pooled_zoom.jpg}
\end{sidewaysfigure}
\clearpage

The reviewer suggests that encoding interference confounded the reading times of all regions after the distractor and argues that this makes the results uninterpretable. The implication seems to be that the patterns observed at the pre-critical, critical, and spillover regions should not be reported.
 However, as we had already reported in the discussion of the SPR results, the two published studies that used the present design \citep{vandyke07, mertzen} show clear indications that these regions of interest were affected by encoding interference as well. Thus, across the three studies (Van Dyke et al., Mertzen et al., and ours) one consistent pattern is that the observed effects could be driven at least partly by encoding interference. We feel that this consistent pattern across studies in the three regions of interest is worth reporting.

### 2024-11-30 paper/revision2/elsarticle-template-harv.tex
minor edits

\begin{abstract}
%% Text of abstract
\Copy{abstract}{
Cue-based retrieval accounts of sentence processing postulate that at a verb, retrieval cues are generated to complete a dependency with the verb's argument(s); for example, the dependency between the subject and the verb must be completed. If these retrieval cues match with not only the \textcolor{blue}{subject} but also with those on other nouns in the sentence, then processing difficulty arises at the verb. This difficulty in identifying the correct dependent is called similarity-based interference. 
We present large-sample self-paced reading and event-related potentials experiments using a well-established design to investigate interference due to syntactic and semantic cues in German. \textcolor{blue}{In this design, the syntactic cue \{+subject\} and the semantic cue \{+animate\} are manipulated.}
\textcolor{blue}{Bayes factors analyses showed} evidence for a semantic interference effect \textcolor{blue}{in both experiments}. 
Surprisingly, Bayes factors provided evidence against interference due to \textcolor{blue}{the syntactic cue \{+grammatical subject\} in this particular design in both experiments}. This finding contradicts the predictions of the standard implementations of cue-based retrieval theory, which (implicitly) assume that both syntactic and semantic cues play an equal role in retrieval. We show through computational modeling that cue-based retrieval \textcolor{blue}{will} also shows no syntactic interference \textcolor{blue}{in the present design} if the parser is assumed to keep track of which clause the subject occurs in. Thus, if syntactic retrieval cues include hierarchical syntactic information (is the noun in the same clause as the verb?), the predictions of the cue-based retrieval model would be consistent with the observed patterns in our data. 
}
\end{abstract}

@@

\subsection*{Discussion}
\Copy{SPR_discussion_revised}{We presented the to-date largest-sample self-paced reading study that \textcolor{blue}{aimed to investigate} the use of syntactic and semantic features during subject-verb dependency formation. \textcolor{blue}{However, reading times started to differ before this critical dependency could be formed. Starting at the distractor which intervened between subject and verb, high semantic interference conditions (animate distractors) were read slower than low semantic interference conditions (inanimate distractors). Because the distractor differed between conditions, we did not analyze distractor reading times. Instead, we treat the pre-critical region which was identical across conditions and preceded the critical verb as a proxy to investigate the difference starting at the distractor. Bayes factors provided extremely strong evidence for a semantic interference effect in the pre-critical region. In contrast, there was evidence against syntactic interference and an interaction in that region. We refrain from interpreting effects in the critical and spill-over region, because it is unclear whether they were caused by the difference starting at the distractor or other processes, i.e., memory retrieval. The reading times differences at the distractor and the pre-critical region are discussed below.}}

@@

\Copy{thisdesign2}{Without the modeling results reported above, it would be easy to conclude that our data show that syntactic cues play no role  \textcolor{blue}{during subject-verb dependency resolution in the investigated sentence configurations}}\label{thisdesign2}. However, this conclusion would be problematic. First, as we showed above, the cue-based retrieval model suggests that -- at least in our data -- the \{$\pm$ grammatical subject\} cue is used, but differently than previously assumed: the relevant cue may be \{$\pm$ subject-in-same-clause\}. Second, there is plenty of independent evidence for the central role of syntactic information in comprehension; some examples are the role of case marking \citep[e.g.,][]{Avetisyan_etal_2020_agree,HusainEtAl2014,bhatia2022processing,bader2000case,bader2006case,miyamoto2002case}, syntactic constraints like Principle A of the binding theory \citep[e.g.,][]{Sturt2003,dillon2013,yadav2021individual},  and 
the importance of word order \citep[e.g.,][]{meng2000mode}.

>>>>

\begin{abstract}
%% Text of abstract
\Copy{abstract}{
Cue-based retrieval accounts of sentence processing postulate that at a verb, retrieval cues are generated to complete a dependency with the verb's argument(s); for example, the dependency between the subject and the verb must be completed. If these retrieval cues match with not only the \textcolor{blue}{subject} but also with those on other nouns in the sentence, then processing difficulty arises at the verb. This difficulty in identifying the correct dependent is called similarity-based interference. 
We present relatively large-sample self-paced reading and event-related potentials experiments using a well-established design to investigate interference due to syntactic and semantic cues in German. \textcolor{blue}{In this design, the syntactic cue \{+subject\} and the semantic cue \{+animate\} are manipulated.}
\textcolor{blue}{Bayes factors analyses showed} evidence for a semantic interference effect \textcolor{blue}{in both experiments}. 
Surprisingly, Bayes factors provided evidence against interference due to \textcolor{blue}{the syntactic cue \{+grammatical subject\} in this particular design in both experiments}. This finding contradicts the predictions of the standard implementations of cue-based retrieval theory, which (implicitly) assumes that both syntactic and semantic cues play an equal role in retrieval. We show through computational modeling that cue-based retrieval \textcolor{blue}{will} also shows no syntactic interference \textcolor{blue}{in the present design} if the parser is assumed to keep track of which clause the subject occurs in. Thus, if syntactic retrieval cues include hierarchical syntactic information (is the noun in the same clause as the verb?), the cue-based retrieval model would exhibit patterns consistent with the observed patterns in our data. 
}
\end{abstract}

@@

\subsection*{Discussion}
\Copy{SPR_discussion_revised}{We presented the to-date largest-sample self-paced reading study that \textcolor{blue}{aimed to investigate} the use of syntactic and semantic features during subject-verb dependency formation. \textcolor{blue}{However, reading times started to differ before this critical dependency could be formed. Starting at the distractor which intervened between subject and verb, high semantic interference conditions (animate distractors) were read slower than low semantic interference conditions (inanimate distractors). Because the distractor differed between conditions, we did not analyze distractor reading times. Instead, to investigate the difference starting after the distractor was read, we investigate the pre-critical region, which was identical across conditions and preceded the critical verb. The pre-critical region serves as a proxy for any encoding interference effects that arose after the distractor was read. Bayes factors provided extremely strong evidence for a semantic interference effect in the pre-critical region. In contrast, there was evidence against syntactic interference and an interaction in that region. We refrain from interpreting effects in the critical and spill-over region, because it is unclear whether they were caused by the difference starting at the distractor or other processes, i.e., memory retrieval. The reading times differences at the distractor and the pre-critical region are discussed below.}}

@@

\Copy{thisdesign2}{Without the modeling results reported above, it would be easy to conclude that, \textcolor{blue}{during subject-verb dependency resolution in the investigated sentence configurations},  our data show that syntactic cues play no role.}\label{thisdesign2} However, this conclusion would be problematic. First, as we showed above, the cue-based retrieval model suggests that -- at least in our data -- the \{$\pm$ grammatical subject\} cue is used, but differently than previously assumed: the relevant cue may be \{$\pm$ subject-in-same-clause\}. Second, there is plenty of independent evidence for the central role of syntactic information in comprehension; some examples are the role of case marking \citep[e.g.,][]{Avetisyan_etal_2020_agree,HusainEtAl2014,bhatia2022processing,bader2000case,bader2006case,miyamoto2002case}, syntactic constraints like Principle A of the binding theory \citep[e.g.,][]{Sturt2003,dillon2013,yadav2021individual},  and 
the importance of word order \citep[e.g.,][]{meng2000mode}.
